---
title: "Spatial Assignment Answers"
author: Zexuan Zhao and Krista Waugh
output: github_document
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(messages = FALSE, cache = FALSE)
```

#Habitat Suitability for Pacific spiny Lumpsucker  

[![Build Status](https://travis-ci.com/espm-157/2018-spatial-spatial_zexuan_krista.svg?token=TJpEEbz7PiUQtcZQe2Sj&branch=master)](https://travis-ci.com/espm-157/2018-spatial-spatial_zexuan_krista)

In this report, we will analysize sea surface temperature, net primary productivity, and marine protected areas along the west coast of the United States to map potential areas of marine aquaculture for the Pacific spiny Lumpsucker (Eumicrotremus orbis). Our conclusion of the study is that the majority of Pacific spiny Lumpsuckers can be found off the coast of the California and Oregon border, with populations spanning the coast of California all the way down to Los Angeles, where there is another large population. 


The following is our spatial data:  

**1. Sea Surface Temperature** (raster data)  
**2. Net Primary Productivity** (raster data)  
**3. Marine Protected Areas**  (vector data)  

```{r libraries, include= FALSE}
library(mapview)      
library(raster)     
library(tidyverse)   
library(sf)          
library(tmap)   
library(spData)
library(leaflet)
```  

## Optimum Pacific spiny Lumpsucker habitat
The temperature range and net primary productivity where the Pacific spiny lumpsucker shows optimum growth is: Sea surface temperatures between 12 and 18 degrees Celsius, and net Primary Productivity between 2.6 and 3 mgC/m2/day. We will begin to look through our data to extract these values to give us a better idea of where, within our dataset, the Pacific spiny lumpsuckers are located. 

### Marine Protected areas 
We'll start with a data file of Marine Protected Areas monitored by the US
Federal government on the west coast: `mpas_westcoast.shp`.
```{r}
westcoast <- st_read("shapefiles/mpas_westcoast.shp")
str(westcoast)
```

### Visual plot of Marine protected areas 
Visually, much of the coastline appears to be managed by the National Marine Fisheries Service, with some influence by National Marine Sanctuaries and Washington and the National Park Service in the north. 

```{r}
westcoast %>% 
  ggplot(aes(fill = State))+
    geom_sf()
```

### Quantitatve analysis of area protected by each marine protection organization
The National Marine Fisheries Service controls by far the largest area of the coastline, 386.5 square kilometers, nearly ten times the area of the next largest organization controlling the Pacific coast of the United States, National Marine Sanctuaries.  

```{r}
area <- westcoast %>% 
  st_area() %>% 
  unclass() 
westcoast %>% 
  mutate(area = area) %>% 
  group_by(State) %>% 
  summarize(area = sum(area)) %>% 
  st_set_geometry(NULL) %>% 
  arrange(desc(area))
```


## Sea Surface Temperature
 We will use the `average_annual_sst_[year].tif` data, which are 5 annual average sea surface temperatures for our region, one for each year fom 2008 to 20012. To make them easier to work with, we will combine them using the stack() function. But first, we will look at the data for each year by plotting them individually, and graphing average sea surface temperature for each year. 
 
```{r sst}
files <- list.files("rasters", full.names = TRUE, pattern = "sst")
sst <- map(files, raster)
names(sst) <- c("sst_2008", "sst_2009","sst_2010", "sst_2011","sst_2012")
sst
```

### Visualizing sea surface temperature from 2008 to 2012
The temperatures are in Kelvin during these steps, but will be converted to Celsius for the final steps. 

```{r}
raster_visualize <- function(raster){
  raster %>% 
    tm_shape() +
    tm_raster() +
    tm_layout(legend.outside = TRUE)
}
sst %>% map(raster_visualize)
```

#### Sea surface temperature average vs maxiumum
2009 had the highest average annual temperature, but the maximum recorded temperature for each year shows that 2011 and 2012 exhibit by far the highest recorded temperatures from our data, with the highest temperature recorded in 2012 being just over 310 degrees Kelvin (37 degrees Celsius), and the lowest maximum temperature in 2010 at about 301 Kelvin (28 degrees celsius). 


```{r warning = FALSE}
data <- data_frame(year = 2008:2012)
for(i in 1:5){
data$value[i] <- sst[[i]] %>% 
    cellStats(mean) 
}
data %>% 
  ggplot(aes(x = year, y = value)) +
    geom_line() +
    ggtitle("Average Annual Sea Surface Temperature")
```

```{r warning = FALSE}
data <- data_frame(year = 2008:2012)
for(i in 1:5){
data$value[i] <- sst[[i]] %>% 
    cellStats(max) 
}
data %>% 
  ggplot(aes(x = year, y = value)) +
    geom_line() + 
    ggtitle("Maximum Recorded Sea Surface Temperature")
```

### Combining the raster layers 
Here we stack our raster layers for each year 2008 to 2012 to get a single layer, which we will then use to get an average sea surface temperature in degrees Celsius.

```{r}
sst_stack <- stack(sst)

```

### Raster calcuations
 
When we stack our rasters, we will write a small custom function to take the mean and convert to Celsius. We will then apply 
this function to our rasterstack to calculate a new raster we can use for analysis.  The conversion between Kelvin
and Celsius is: 
$C = K - 273.15$



```{r}
convert <- function(x){
  x - 273.15
}
raster_mean <- function(raster){
  cellStats(raster, mean) %>% 
    convert()
}
```


```{r}
sst_stack_celsius <- calc(sst_stack, fun = convert)
plot(sst_stack_celsius)
```

## Net Primary Productivity 
Since Lumpsuckers may be influenced by more than just sea surface temperature, 
we also include Net Primary Production (NPP) in our analysis. 

#### NPP raster data   

We used the raster() command to read in our data for net primary productivity along the west coast of the US. 

```{r}
npp <- raster("rasters/annual_npp.tif")
plot(npp)
```

### Reprojecting net primary productivity and sea surface temperature 
Here we reset the CRS of sea surface temperature to match the CRS of net primary productivity. 

 

```{r warning = FALSE}
sst_stack_projected <- projectRaster(from = sst_stack_celsius, to = npp)
summary(values(sst_stack_projected))
```


### Finally, we can finish our data preparation and begin our analysis

```{r}
sst_npp <- stack(sst_stack_projected, npp)
```


## Analysis   

Now that our data is prepped, we can move onto the analysis. For this specific analysis, we use 
the SST and NPP data to find areas along the US West Coast that are suitable for 
growing lumpsucker fish. This requires removal of all cells from NPP and SST 
that are not within the ideal growth parameter range.

### Sample Points & Extract values from Rasters
We generate 1000 sample points from the Pacific coast of the US in order to represent sea surface temperature and net primary productivity of our area of study. 

```{r}
set.seed(2018)
random_points <- st_sample(westcoast, 1000, type = "random") %>%
  st_sf() %>% 
  na.omit() %>% 
  st_join(westcoast)
plot(st_geometry(random_points))
```

#### R Question: Why does your new dataframe of points likely have fewer than 1000 points? 
See the `st_sample()` documentation and explain. 

### Extract Raster Values and Identify Habitat
Here, we extracted values from the combined data of sea surface temperature and net primary productvity to identify Pacific spiny Lumpsuckers optimum habitat areas. Lumpsucker fish grow best in waters that are **between 12 and 18 degrees Celsius.** and with an NPP between **2.6 and 3 mgC/m2/day**.


```{r}
sample_points <- raster::extract(sst_npp, random_points, sp = T) %>% 
  na.omit() %>% 
  st_as_sf() %>% 
  select(starts_with("sst"), ends_with("npp") ) %>% 
  mutate(point_ID = 1:nrow(.)) %>% 
  gather(year, sst, starts_with("sst")) %>% 
  separate(year, into = c("key", "year"), sep = "_") %>% 
  select(point_ID, year, annual_npp, sst) %>% 
  group_by(point_ID) %>% 
  summarize(habitat = (max(sst) < 18 & min(sst) > 12 & mean(annual_npp) > 2.6 & mean(annual_npp) < 3))
sample_points %>% 
  ggplot(aes(col = habitat)) +
  geom_sf()
```

### Percentage of sampled points where Lumpsuckers would be found 
```{r}
sample_points %>% 
  summarize(percentage = sum(habitat, na.rm = TRUE) / n()) %>% 
  st_set_geometry(NULL) %>% 
  pull()
```

#### Minimum latitude where lumpsucker fish are likely to be found
```{r}
coordinates <- sample_points %>% 
  filter(habitat == TRUE) %>% 
  st_geometry() %>% 
  st_transform(4236) %>% 
  st_coordinates() %>% 
  as_data_frame() %>% 
  summarize(max_long = max(X),
            min_long = min(X),
            max_lat  = max(Y),
            min_lat  = min(Y))
coordinates$min_lat
```

#### Pacific spiny Lumpsucker habitat  
```{r}
sample_points %>% 
  filter(habitat == TRUE) %>% 
  st_geometry() %>% 
  st_transform(4236) %>%
  ggplot() +
    geom_sf() +
    geom_sf(data = st_geometry(world)) +
    coord_sf(xlim = c(coordinates$min_long,
                      coordinates$max_long),
             ylim = c(coordinates$min_lat,
                      coordinates$max_lat)) 
```

The majority of Pacific spiny Lumpsuckers can be found off the coast of the California and Oregon border, with populations spanning the coast of California all the way down to Los Angeles, where there is another large population. See the leaflet below:  

```{r}
leaflet() %>% 
  setView(lng = -123.5, lat = 38.5, zoom = 5) %>%
  addTiles() %>% 
  addMarkers(-127, 43, label = "Habitat 1") %>% 
  addMarkers(-119, 33, label = "Habitat 2")
```

